---
title: "Mixed-Effects Models in BRMS"
author: "Martin Zettersten"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

# Preliminaries

Install packages if they aren't already installed (notice the nice compact code here that avoids re-installing the package if it already is found in your R environment).

```{r}
if (!("knitr" %in% installed.packages())) install.packages("knitr")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("here" %in% installed.packages())) install.packages("here")
if (!("brms" %in% installed.packages())) install.packages("brms")
if (!("broom.mixed" %in% installed.packages())) install.packages("broom.mixed")
if (!("tidybayes" %in% installed.packages())) install.packages("tidybayes")
if (!("bayesplot" %in% installed.packages())) install.packages("bayesplot")
if (!("modelr" %in% installed.packages())) install.packages("modelr")
```

Load libraries and set up paths

```{r warning=F,message=F}
library(knitr)
library(brms)
library(tidyverse)
library(here)
library(tidybayes)
library(broom.mixed)
library(bayesplot)
library(modelr)

data_path <- here("data")
```

# General takeaway

An elegant aspect of R's infrastructure is that we can also easily adapt our analyses to a Bayesian regression approach. Here we'll use the "brms" package to fit a Bayesian linear mixed-effects model to the same data we used in the introduction to linear mixed-effects models. Check out the brms documentation for many more vignettes and examples here: <https://paulbuerkner.com/brms/index.html>

# Inspect the data

Load data, and name that object "rt_data"

```{r}
rt_data <- read_csv(here(data_path,"rt_dummy_data.csv"))
```

# Recode variables and transform DV

```{r}
rt_data <- rt_data %>%
  mutate(
    #modality_c = ifelse(modality=="Audio-only",-0.5,0.5),
    log_rt = log(RT)
  )
```

# Fit a Bayesian linear mixed-effects model

Let's first fit a Bayesian regression model predicting log reaction time from modality, while accounting for the nested structure of the data, while leaving all of the model parameters at their defaults.

## Fit a default brms model

Notice that the core model syntax is *exactly the same* as for lme4.

```{r, cache=TRUE, warning=FALSE, message=FALSE}
#this takes a while to fit - note the use of caching to save time on future iterations
rt_mod_log <- brm(log_rt ~ 1 + modality + 
                      (1 + modality|PID),
                    data = rt_data)
```

### Summary

Inspect the summary output.

```{r}
summary(rt_mod_log)
```

Summarize specific coefficients using `tidybayes`.

```{r}
rt_mod_log %>%
  spread_draws(b_modalityAudiovisual) %>%
  median_qi()
```


### Posterior Predictive Checks

Check the posterior predictive distribution, i.e. compare the observed data (y) to simulated data from the posterior predictive distribution (y_rep).
If the model fits the data well, then data we generate (simulate) from the model should look similar to the raw data observed.

```{r}
pp_check(rt_mod_log)
```

### Plot MCMC Samples

```{r}
#plot(rt_mod_log)
```

### Plot the modality effect

```{r}
#plot estimated means for each condition tidybayes
plot(conditional_effects(rt_mod_log))
```

### Plot the posterior distributions

```{r}
posterior <- as.matrix(rt_mod_log)

bayesplot::mcmc_areas(posterior,
           pars = c("b_modalityAudiovisual"),
           # arbitrary threshold for shading probability mass
           prob = 0.83) 
```

### Prettier

```{r}
post_data <- rt_data %>% 
  data_grid(modality) %>%
  add_epred_draws(rt_mod_log, ndraws = 4000, re_formula = NA)

#aggregate participant data
rt_data_agg <- rt_data %>% 
  group_by(PID, modality) %>% 
  summarise(avg_rt = mean(log_rt), .groups = "drop")

post_data %>%
  ggplot(aes(x=modality,y=.epred,fill=modality))+
  #geom_jitter(data=rt_data_agg,aes(y=avg_rt,color=modality),position = position_nudge(x = -0.1),alpha=0.5,stroke=NA)+
  stat_halfeye()+
  ylab("Log Reaction Time (ms)")+
  xlab("Modality")+
  theme_minimal()+
  theme(legend.position="none")
```

## Specify model structure more explicitly

Second, Bayesian models are extremely flexible, but that means we also need to specify many aspects of our model more explicitly. 
Two key features of the model to specify are the model family and the priors.
Can you figure out what family was used in the above example?

```{r}
#extract the family
family(rt_mod_log)
#extract the priors
prior_summary(rt_mod_log)
```

The model family is specified with the `family` argument. 
Here, we'll now set the Gaussian family explicitly (i.e., assuming a normal distribution), which is the default for continuous outcomes.

Next, we show an example of how to set the model parameters (coefficients, SDs and covariances) more explicitly.

### Specify the prior

```{r}
#set the prior
priors <-c(
  set_prior("normal(6.9, 1)", class = "Intercept"), #normal distribution for intercept around ~ log(1000) ms (just for illustration)
  set_prior("normal(0, 1)", class = "b"), # normal distribution for fixed-effect coefficients
  set_prior("normal(0, 0.5)", class = "sd"), #normal distribution for sd/ random effects
  set_prior("lkj(2)", class = "L")) # lkj distribution for covariance matrix
```

### Fit the model with family and prior more explicit

```{r, cache=TRUE, warning=FALSE, message=FALSE}
#this takes a while to fit - note the use of caching to save time on future iterations
rt_mod_log_w_priors <- brm(log_rt ~ 1 + modality + 
                      (1 + modality|PID),
                    data = rt_data,
                  prior=priors, #our specified priors
                  family=gaussian() #could also set log as link function
                  )
summary(rt_mod_log_w_priors)
```

### Plot

```{r}
post_data_w_priors <- rt_data %>% 
  data_grid(modality) %>%
  add_epred_draws(rt_mod_log_w_priors, ndraws = 4000, re_formula = NA)

post_data_w_priors %>%
  ggplot(aes(x=modality,y=.epred,fill=modality))+
  stat_halfeye()+
  ylab("Log Reaction Time (ms)")+
  xlab("Modality")+
  theme_minimal()+
  theme(legend.position="none")
```

# Session Info

```{r}
sessionInfo()
```
